\section{Chapter 3}

We will make clear some asymptotic notation used in the text.
\begin{definition}
	\label{def:big-theta}
	Given functions $f,g:\mathbb{N}\to\mathbb{R}$ we say $f(n) = \Theta(g(n))$ if there exists $c_{1},c_{2}>0$ and $n_{0}\in\mathbb{N}$ such that $c_{2}\left|g(n)\right|\leq \left|f(n)\right|\leq c_{1}\left|g(n)\right|$. More generally, we consider $\Theta(g(n))$ to be the set of all functions that satisfy the above statement.
\end{definition}

\begin{definition}
	\label{def:big-O}
	Given functions $f,g:\mathbb{N}\to\mathbb{R}$ we say $f(n)=O(g(n))$ if there exists $c>0$ and $n_{0}\in\mathbb{N}$ such that $\left|f(n)\right|\leq c\left|g(n)\right|$ for all $n\geq n_{0}$. More generally, we consider $O(g(n))$ to be the set of all functions that satisfy the above statement.
\end{definition}

\begin{definition}
	\label{def:big-omega}
	Given functions $f,g:\mathbb{N}\to\mathbb{R}$ we say $f(n)=\Omega(g(n))$ if there exists $c>0$ and $n_{0}\in\mathbb{N}$ such that $c\left|g(n)\right|\leq\left|f(n)\right|$ for all $n\geq n_{0}$. More generally, we consider $\Omega(g(n))$ to be the set of all functions that satisfy the above statement.
\end{definition}

Notice that in all three of the above definitions we can consider the defined structure as a set or a identity between two functions. In this case we can write either $f\in O(g(n))$ or $f=O(g(n))$. Arguably the correct notation is $f\in O(g(n))$ as this works for the most general case where we consider $O(g(n))$ to be a set. However, we will often use the equality notation as a matter of simplicity since the function we compare with remains constant and we wish to make a statement about a particular element of the set in question.

We have the following two definitions which are more restrictive than those above.
\begin{definition}
	\label{def:little-oh}
	Given functions $f,g:\mathbb{N}\to\mathbb{R}$ we say $f(n)=o(g(n))$ if for every $c>0$ there exists $n_{0}\in\mathbb{N}$ such that $\left|f(n)\right|<c \left|g(n)\right|$ for all $n\geq n_{0}$. More generally, we consider the set $o(g(n))$ to be the set of all functions that satisfy the above statement.
\end{definition}

\begin{definition}
	\label{def:little-omega}
	Given functions $f,g:\mathbb{N}\to\mathbb{R}$ we say $f(n)=\omega(g(n))$ if for any positive constant $c>0$ there exists $n_{0}\in\mathbb{N}$ such that $c\left|g(n)\right|<\left|f(n)\right|$ for all $n\geq n_{0}$. More generally, we consider $\omega(g(n))$ to be the set of all functions that satisfy the above statement.
\end{definition}

\subsection{Asymptotic Notation - Exercises}

\subsubsection*{3.1--1}

Suppose $f,g:\mathbb{N}\to\mathbb{R}$ are asymptotically non-negative. Then there exists $n_{f}, n_{g}\in\mathbb{N}$ such that,
\begin{equation*}
	\begin{aligned}
		0 \leq f(n)& &\forall n\geq n_{f}\\
		0 \leq g(n)& &\forall n\geq n_{g}
	\end{aligned}
\end{equation*}
If we take $n_{0}:=\max(n_{f}, n_{g})$ then,
\begin{equation*}
	0\leq f(n), g(n) \quad \forall n\geq n_{0}
\end{equation*}
and so,
\begin{equation*}
	0 \leq \max(f(n), g(n)) \quad \forall n\geq n_{0}
\end{equation*}
By definition,
\begin{equation*}
	\max(f(n),g(n)) =
	\begin{cases}
		f(n) & \text{if }f(n)\geq g(n)\\
		g(n) & \text{otherwise}
	\end{cases}
\end{equation*}
so by the asymptotic non-negativity, $\max(f(n), g(n))\leq f(n) + g(n)$ and $\max(f(n), g(n))\geq \frac{f(n)+g(n)}{2}$ for all $n\geq n_{0}$. Hence, $\max(f(n), g(n))\in\Theta(f(n)+g(n))$.

\subsubsection*{3.1--2}

Let $a,b\in\mathbb{R}$ such that $b>0$. Choose $n_{0}\in\mathbb{N}$ such that $n_{0}\geq\lceil|a|\rceil$ then,
\begin{equation*}
	(n+a)^{b}\leq 2^{b}n^{b} \quad \forall n\geq n_{0}
\end{equation*}
so $(n+a)^{b}=O(n^{b})$. Working in reverse,
\begin{equation*}
	\begin{aligned}
		(n+a)^{b} &\geq cn^{b}\\
		n+a &\geq c^{\frac{1}{b}}n\\
		n-c^{\frac{1}{b}}n &\geq -a\\
		n &\geq \frac{-a}{1-c^{\frac{1}{b}}}
	\end{aligned}
\end{equation*}
it then suffices to choose $c<1$. So for any $c=\frac{1}{2}$ and $n_{0}\geq \frac{-a}{1-c^{\frac{1}{b}}}$ we then have,
\begin{equation*}
	(n+a)^{b}\geq cn^{b}
\end{equation*}
hence, $(n+a)^{b}\in\Omega(n^{b})$. Therefore, $(n+a)^{b}\in\Theta(n^{b})$.

\subsubsection*{3.1--3}

The statement ``The running time of the algorithm $A$ is at least $O(n^{2})$'' is meaningless as $O(n^{2})$ gives an asymptotic upper bound, or in terms of algorithms, the worst-case running time. Since $f(n)=O(n^{2})$ means there exists $c>0$ and $n_{0}\in\mathbb{N}$ such that,
\begin{equation*}
	0\leq f(n)\leq cn^{2}\quad\forall n\geq n_{0}
\end{equation*}
It is clear from the definition that the algorithm runtime may or may not be less than $cn^{2}$.

\subsubsection*{3.1--4}

\begin{itemize}
	\item [(a)]
		Given $c\geq 2$ we can write $2^{n+1}=2\cdot2^{n}\leq c2^{n}$ for any $n\in\mathbb{N}$. So $2^{n+1}=O(2^{n})$.
	\item [(b)]
		Suppose there exists $c>0$ and $n_{0}\in\mathbb{N}$ such that $0\leq 2^{2n}\leq c2^{n}$ for all $n\geq n_{0}$.Then $0\leq 2^{n} \leq c$ for all $n\geq n_{0}$ which is clearly a contradiction. Hence, $2^{2n}\neq O(2^{n})$.
\end{itemize}

\subsubsection*{3.1--5}

\begin{theorem}
	\label{thm:big-theta}
	For any two functions $f,g:\mathbb{N}\to\mathbb{R}$ we have $f(n)=\Theta(g(n))$ if and only if $f(n)=O(g(n))$ and $f(n)=\Omega(g(n))$.
\end{theorem}

\begin{proof}
	First suppose $f(n)=\Theta(g(n))$, then by definition there exists $c_{1},c_{2}>0$ and $n_{0}\in\mathbb{N}$ such that,
	\begin{equation*}
		c_{2}\left|g(n)\right| \leq \left|f(n)\right| \leq c_{1}\left|g(n)\right| \quad\forall n\geq n_{0}
	\end{equation*}
	and so $f(n)$ satisfies the conditions of $O(g(n))$ and $\Omega(g(n))$.

	Now consider $f(n)=O(g(n))$ and $f(n)=\Omega(g(n))$. Then, there exists $c_{1},c_{2}>0$ and $n_{1},n_{2}\in\mathbb{N}$ such that,
	\begin{equation*}
		\begin{aligned}
			\left|f(n)\right| &\leq c_{1}\left|g(n)\right|& &\forall n\geq n_{1}\\
			c_{2}\left|g(n)\right| &\leq \left|f(n)\right|& &\forall n\geq n_{2}
		\end{aligned}
	\end{equation*}
	Set $n_{0}:=\max(n_{1},n_{2})$ then we have,
	\begin{equation*}
		c_{2}\left|g(n)\right| \leq \left|f(n)\right| \leq c_{1}\left|g(n)\right|\quad\forall n\geq n_{0}
	\end{equation*}
	and so $f(n)=\Theta(g(n))$.
\end{proof}

\subsubsection*{3.1--6}

\begin{corollary}
	\label{cor:big-theta}
	The running time of an algorithm is $\Theta(g(n))$ if and only if its worst-case running time is $O(g(n))$ and its best-case running time is $\Omega(g(n))$.
\end{corollary}

\begin{proof}
	This is an immediate application of Theorem \ref{thm:big-theta}. Let $W(n)$ denote the worst-case running time, $B(n)$ denote the best-case running time and $T(n)$ denote the running time of the algorithm. Then,
	\begin{equation*}
		B(n) \leq T(n) \leq W(n)\quad\forall n\in\mathbb{N}
	\end{equation*}
	and so $T(n)=O(g(n))$ and $T(n)=\Omega(g(n))$. Hence, by Theorem \ref{thm:big-theta} $T(n)=\Theta(g(n))$.
\end{proof}

\subsubsection*{3.1--7}

\begin{proposition}
	\label{prop:little-intersection}
	The set $o(g(n))\cap\omega(g(n))$ is empty.
\end{proposition}

\begin{proof}
	It is sufficient to show that for any $f(n)=o(g(n))$ we have $f(n)\neq \omega(g(n))$. Suppose $f(n)=o(g(n))$ then for every $c>0$ there exists $n_{0}\in\mathbb{N}$ such that $\left|f(n)\right| < c\left|g(n)\right|$ for all $n\geq n_{0}$. Assume $f(n)=\omega(g(n))$ then for any $c>0$ we can choose $n_{0}\in\mathbb{N}$ so that,
	\begin{equation*}
		c\left|g(n)\right| < \left|f(n)\right| < c\left|g(n)\right|
	\end{equation*}
	This is clearly a contradiction and so we have $f(n)\neq \omega(g(n))$. Hence, $o(g(n))\cap\omega(g(n))=\emptyset$.
\end{proof}

\subsubsection*{3.1--8}

\begin{equation*}
	\begin{aligned}
		O(g(n,m)) := \{f(n,m)\colon&\exists c>0 \text{ and } \exists n_{0},m_{0}\in\mathbb{N} \text{ such that }\\
		&\left|f(n,m)\right| \leq c\left|g(n,m)\right|\,\forall n\geq n_{0}\text{ or }m\geq m_{0}\}\\
		\Omega(g(n,m)) := \{f(n,m)\colon&\exists c>0 \text{ and } \exists n_{0},m_{0}\in\mathbb{N} \text{ such that }\\
		&c\left|g(n,m)\right| \leq \left|f(n,m)\right|\,\forall n\geq n_{0}\text{ or }m\geq m_{0}\}\\
		\Theta(g(n,m)) := \{f(n,m)\colon&\exists c_{1},c_{2}>0 \text{ and } \exists n_{0},m_{0}\in\mathbb{N} \text{ such that }\\
		&c_{2}\left|g(n,m)\right|\leq\left|f(n,m)\right| \leq c_{1}\left|g(n,m)\right|\,\forall n\geq n_{0}\text{ or }m\geq m_{0}\}
	\end{aligned}
\end{equation*}
